sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-09-08 doc/paper-jss/paper-jss.Rmd
Overhauling examples

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
    address: |
      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
    address: |
      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{pdflscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \usepackage{gensymb}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
  \newunicodechar{˚}{\degree}
  \DeclareGraphicsExtensions{.pdf,.png,.jpg}
  \widowpenalty10000
  \clubpenalty10000
output: rticles::jss_article
---

@@

# Introductory example using fish survey data

## Illustrating the need for spatial random fields

We begin with a simple species distribution model of encounter probability of
Pacific Cod (TODO spp) from a trawl survey conducted off the west
coast of Vancouver Island, British Columbia, Canada. This survey consists of spatially random sampling in each year, resulting in a georeferenced dataset. For compatibility with other packages, spatial random fields in \pkg{sdmTMB} can be toggled on and off. We will fit a logistic
regression of encounter probability with and without spatial random fields to
illustrate the importance of accounting for spatial correlation. In addition to the spatial random field, we include an intercept and a quadratic effect of depth on the probability of occurrence.

@@

We can create a 'mesh' for the SPDE calculations with `make_mesh()`.

@@

We will fit a model with and without spatial random fields by toggling the `spatial` argument from `"on"` to `"off`. The `update()` S3 method refits the same model updating any specified arguments.

```{r}
fit_bin_rf <- sdmTMB(
  present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, 
  spatial = "on",
  family = binomial(link = "logit")
)
fit_bin <- update(fit_bin_rf, spatial = "off")
```

We can get a summary of our model fit by printing it.

```{r dog-bin-summary, echo=TRUE}
fit_bin_rf
```

Describe what we see here

We can test for spatial autocorrelation with a Moran's I test or with a visual inspection of the residuals.

@@

```{r, echo=FALSE, results='hide'}
test_autocor <- function(obj) {
  set.seed(1)
  s <- simulate(obj, nsim = 500)
  pr <- predict(obj, type = "response")$est
  r <- DHARMa::createDHARMa(
    simulatedResponse = s,
    observedResponse = pcod$present,
    fittedPredictedResponse = pr
  )
  DHARMa::testSpatialAutocorrelation(r, x = pcod$X, y = pcod$Y, plot = FALSE)
}
test_autocor(fit_bin)
test_autocor(fit_bin_rf)
```

@@

```{r}
AIC(fit_bin_rf, fit_bin)
```

We could do additional testing with cross validation via `sdmTMB_cv()`. Here we will do 10-fold cross validation with the folds constructed randomly. Using the `fold_ids` argument, we could supply our own folds and conduct spatially blocked cross validation, for example.

```{r}
cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, spatial = "on",
  family = binomial(link = "logit"), k_folds = 10
)
cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, spatial = "off",
  family = binomial(link = "logit"), k_folds = 10
)
```

@@

```{r}
cv_bin_rf$sum_loglik
cv_bin_rf$sum_loglik
```

The standard errors on our fixed effects have increased:

```{r}
tidy(fit_bin_rf, conf.int = T)
tidy(fit_bin, conf.int = T)
```

We now have an effect of depth and a spatial random field, which combined in link space result in our overall prediction. Note the difference between only the fixed effects (left) and including the spatial random field (right).

```{r}
p <- predict(fit_bin_rf, newdata = qcs_grid)
plot_spatial_map <- function(dat, column, title) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    coord_fixed() +
    theme(legend.position = "bottom") +
    ggtitle(title)
}
g1 <- plot_spatial_map(p, plogis(est_non_rf), "Fixed effects")
g2 <- plot_spatial_map(p, omega_s, "Spatial random field") +  scale_fill_gradient2()
g3 <- plot_spatial_map(p, plogis(est), "Combined prediction")
cowplot::plot_grid(g1, g2, g3, ncol = 3)
```

@@

As a second example, we will construct a model of catch rates of North Pacific spiny dogfish from the same survey. This example extends the spatial model described previously by also including (1) spatiotemporal fields allowing the distribution in each year to be unique (modelled as an autoregressive process, allowing hotspots to persist through time), (2) a time varying intercept to allow each year to have a different mean catch rate, and (3) a smooth effect of ln(depth), allowing catch rates to vary non-linearly with depth. As catch rates are both skewed and zero inflated, we specify the response family as a Tweedie distribution [@tweedie1984].

```{r dog-head-dat, echo=TRUE}
dat <- select(dogfish, longitude, latitude, year, catch_weight, area_swept, depth)
head(dat)
```

```{r dog-utms, echo=TRUE}
dat <- add_utm_columns(dat, c("longitude", "latitude"), units = "km", utm_crs = 32609)
dat$log_depth <- log(dat$depth)
```

```{r dog-mesh, echo=TRUE}
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
```

@@

```{r dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
fit_tw <- sdmTMB(
  catch_weight ~ s(log_depth),
  data = dat,
  mesh = mesh,
  offset = log(dat$area_swept),
  time = "year",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  spatial = "on",
  spatiotemporal = "iid",
  anisotropy = TRUE,
  share_range = TRUE,
  silent = FALSE,
  family = tweedie()
)
```

We can use `update()` to refit our model quickly without specifying all of the arguments. This is highlighted by exploring models with four additional families, each using a delta- or hurdle distribution:
delta-gamma, delta-lognormal, delta-Poisson-link-gamma, and
delta-Poisson-link-lognormal.

@@

We can then compare the fits via AIC:

@@

We find that the Poisson-link-lognormal model is favoured with AIC. We next
test two additional model formulations: making the spatial correlation
isotropic (the default) instead of anisotropic and structuring the positive model component
spatiotemporal random fields as AR(1).

@@

<!-- Need description of anisotropy and why it's important here or somewhere else -->
We find that the anisotropic AR(1) is favoured. On inspecting the model output
with `summary()`, we notice that the spatial random field SD in the positive model component has become very
small once AR(1) spatiotemporal fields were added. We therefore consider
a model where these are turned off in the second component of the delta model:

@@

@@

```{r dog-aniso, echo=TRUE}
plot_anisotropy(fit_dpl)
```

<!-- Propose moving this to a short paragraph `Prediction` to highlight modularity -->
```{r dog-grid, echo=TRUE}
head(wcvi_grid, n = 3)
grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
grid$log_depth <- log(grid$depth)
head(grid, n = 3)
```

@@

Our prediction data frame has..., which we can easily plot with ggplot2 or similar. We suppress that code in this paper for brevity.

@@

@@

```{r dog-wvci-pred, fig.asp = 0.66, fig.cap="Example prediction components..."}
cowplot::plot_grid(
  g_nonrf,
  g_omega,
  g_eps,
  g_est,
  ncol = 2L
)
```

```{r dog-depth, echo=TRUE}
nd <- data.frame(
  log_depth = seq(min(dat$log_depth), max(dat$log_depth), length.out = 100), 
  year = max(dat$year)
)
pred_depth <- predict(fit, newdata = nd, se_fit = TRUE, re_form = NA, model = NA)
```

```{r dog-depth-plot, echo=FALSE}
ggplot(pred_depth, aes(
  exp(log_depth), exp(est), 
  ymin = exp(est - 2 * est_se), 
  ymax = exp(est + 2 * est_se))) +
  geom_ribbon(fill = "grey90") +
  geom_line() +
  labs(x = "Depth (m)", y = "Density")
```

@@

```{r dog-index-plot, echo=TRUE}
ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
  geom_pointrange() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(y = "Biomass", x = "Year")
```

@@

@@

The SVC effect at any point in space is the combination of the fixed effect and spatial random effect for `nao`. Here we illustrate

We can extract the SVC effect with TODO

```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
nsim <- 200
zeta_s <- predict(fit_owl, newdata = snow, nsim = nsim, sims_var = "zeta_s")
sims <- spread_sims(fit_owl, nsim = nsim)
combined <- sims$nao + t(zeta_s)
snow$nao_effect <- apply(exp(combined), 2, median)
```

We can make a basic plot using the following code. A more elaborate version is shown in Figure \ref(fig:owl-plot-fancy).

```{r owl-plot-basic, echo=TRUE, eval=FALSE}
ggplot(snow, aes(X, Y)) +
  geom_point(aes(colour = nao_effect)) +
  scale_colour_viridis_c()
```

Overall, we found a weak average positive effect of annual mean NAO on overall counts, but a southeast to northwest gradient in the intensity of the effect (Figure \ref{fig:owl-nao}).
This result is consistent with owls closest to the Atlantic coast and those migrating the furthest south being the most affected by NAO.
We provide an extended version of this example in Appendix \ref{app:owls}.

@@

Importantly, `dependencies = TRUE` includes the \proglang{R}-INLA package  [@rue2009; @lindgren2011; @lindgren2015], which is not on CRAN but is needed for the `make_mesh()` function below.

>>>>

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
    address: |
      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
    address: |
      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that \pkg{sdmTMB} can be an order of magnitude faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \usepackage{gensymb}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
  \newunicodechar{˚}{\degree}
  \DeclareGraphicsExtensions{.pdf,.png,.jpg}
  \widowpenalty10000
  \clubpenalty10000
output: rticles::jss_article
---

@@

# Installation

Importantly, `dependencies = TRUE` includes the \proglang{R}-INLA package
[@rue2009; @lindgren2011; @lindgren2015], which is not on CRAN but is needed
for the `make_mesh()` function below. Soon, we intend to switch to using the
\pkg{fmesher} package to eliminate this dependency.

Alternatively, the development version can be installed with:

```{r remotes, eval=FALSE, echo=TRUE}
# install.packages("remotes")
remotes::install_github("pbs-assess/sdmTMB", dependencies = TRUE)
```

Additional utilities are maintained in the \pkg{sdmTMBextra} package at
<https://github.com/pbs-assess/sdmTMBextra>.

# Species distribution modelling with spatial random fields

We begin with a simple species distribution model of encounter probability of
Pacific Cod (*Gadus macrocephalus*) from a trawl survey conducted in Queen
Charlotte Sound, British Columbia, Canada. The purpose of our example is to
illustrate the need for spatial random fields. This survey conducts
depth-stratified random sampling biennially, resulting in a georeferenced
dataset. The dataset includes columns for year, latitude, longitude, Universal
Transverse Mercator (UTM) zone 9 coordinates, bottom depth, and encounter vs.
non-encounter.

@@

```{r pcod-head, echo=TRUE}
select(pcod, year, lat, lon, X, Y, depth, present) |> 
  head(n = 3)
```

An \pkg{sdmTMB} model requires a data frame that contains a response column,
columns for any predictors, and columns for spatial coordinates.
Usually it makes sense to convert the spatial coordinates to an equidistant
projection such as the Universal Transverse Mercator (UTM) to ensure that
distance remains constant throughout the study region [e.g., using
`sf::st_transform()`, @pebesma2018]. Here we use the helper function
`add_utm_columns()` to add UTM coordinates with km units (so our estimated
spatial range parameter is not too big or small). By default, the function will
guess the UTM zone and create new columns `X` and `Y`. Since our example
data set already has these UTM columns, we can skip running this code.

We then create a mesh object that contains triangulation and projection
matrices needed to apply the SPDE approach. Here, `cutoff` defines the minimum
allowed distance between mesh vertices in the units of `X` and `Y`.
`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()` and `cutoff`
is passed through to it. Soon, \pkg{sdmTMB} will switch to using the
\pkg{fmesher} package for this purpose, which is hosted on CRAN. Alternatively,
we could have created any mesh via \proglang{R}-\pkg{INLA} or \pkg{fmesher} and
supplied it to the `mesh` argument in `make_mesh()`. We can inspect our mesh
object with the associated plotting method.

@@

We can then fit our model with `sdmTMB()`. We will fit a logistic regression of
encounter probability with and without spatial random fields (by toggling the
`spatial` argument) to illustrate the importance of accounting for spatial
correlation. In addition to the spatial random field, we include an intercept
and a quadratic effect of depth on the probability of encounter. The `update()`
S3 method refits the same model updating any specified arguments.

Our random field model can be written as TODO.

```{r, echo=TRUE}
fit_bin_rf <- sdmTMB(
  present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, 
  spatial = "on",
  family = binomial(link = "logit")
)
fit_bin <- update(fit_bin_rf, spatial = "off")
```

```{r pcod-eg1-sanity, eval=TRUE, echo=TRUE, message=FALSE}
sanity(fit)
```

Here, this does not flag any issues. `sanity()` is checking that the `nlminb()`
optimizer reported successful convergence, that the Hessian matrix is positive
definite, that no extreme (XX) or small (XX) eigenvalues are detected, that no
absolute log likelihood gradients with respect two fixed defects are >= 0.001,
that all fixed effects have reported standard errors and that they do not look
unreasonably large (XX by default), that random field marginal standard
deviations are not unexpectedly small or large, and that the random field
Matérn range parameter does not look unreasonably large.

We can get a summary of our model fit:

```{r pcod-bin-summary, echo=TRUE}
summary(fit_bin_rf)
```

The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
We also see the formula, mesh, fitted data, and family. Next we see any
estimated main effects, the Matérn range distance, the spatial random field
standard deviation, and the negative log likelihood at convergence.

<!-- The -->
<!-- smoother summary follows the format used in the \pkg{brms} package -->
<!-- [@burkner2017]. As noted in the print method, we can extract a tidy data frame -->
<!-- similar to the format in the \pkg{broom} [@broom] and \pkg{broom.mixed} -->
<!-- [@broommixed] packages: -->

@@

```{r, echo=FALSE, results='hide'}
test_autocor <- function(obj) {
  set.seed(1)
  s <- simulate(obj, nsim = 500)
  pr <- predict(obj, type = "response")$est
  r <- DHARMa::createDHARMa(
    simulatedResponse = s,
    observedResponse = pcod$present,
    fittedPredictedResponse = pr
  )
  DHARMa::testSpatialAutocorrelation(r, x = pcod$X, y = pcod$Y, plot = FALSE)
}
(t_no_rf <- test_autocor(fit_bin))
(t_rf <- test_autocor(fit_bin_rf))
p_rf <- round(t_rf$p.value, 2)
```

We can test for spatial autocorrelation with a Moran's I test or with a visual
inspection of the residuals calculated from the `residuals()` S3 method. We
omit the Moran's I code for brevity but include it in the code supplement. The
p-value for spatial autocorrelation is low for the model without a random field
(p < 0.01) but large for the model with a random field (p = `r p_rf`)
suggesting significant residual correlation in the simpler model that is
alleviated by including the random field.

@@

```{r pcod-aic, eval=TRUE, echo=TRUE, results='markup'}
AIC(fit_bin_rf, fit_bin)
```

We could do additional testing with cross validation via `sdmTMB_cv()`. Here we
will do 10-fold cross validation with the folds constructed randomly. Using the
`fold_ids` argument, we could supply our own folds and conduct spatially
blocked cross validation.

```{r pcod-cv, eval=TRUE, echo=TRUE}
cv_bin_rf <- sdmTMB_cv(present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, spatial = "on",
  family = binomial(), k_folds = 10
)
cv_bin <- sdmTMB_cv(present ~ poly(log(depth), 2),
  data = pcod, mesh = mesh_pcod, spatial = "off",
  family = binomial(), k_folds = 10
)
```

@@

```{r pcod-cv-out, eval=TRUE, echo=TRUE, results='markup'}
cv_bin_rf$sum_loglik
cv_bin$sum_loglik
```

The standard errors on our fixed effects have increased with the spatial random
field:

```{r pcod-tidy, eval=TRUE, echo=TRUE, results='markup'}
tidy(fit_bin_rf, conf.int = TRUE)
tidy(fit_bin, conf.int = TRUE)
```

\sloppy
To visualize our model, we can make predictions with the `predict()` method
(`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on
a new data frame with any locations and values for the predictor columns. Here,
we will predict on a 2 $\times$ 2 km grid (`qcs_grid`) that covers the entire
region of interest so we can visualize the predictions spatially. The grid
contains spatial covariate columns and all predictors used in the model set at
values for which we want to predict. The output of `predict()` is a data frame
containing overall estimates in link space (`est`), estimates from the
non-random-field components (`est_non_rf`; here, intercept and depth), and
estimates from the individual random field components (here, `omega_s`---the
spatial field).

```{r pcod-predict, echo=TRUE, results='markup'}
p <- predict(fit_bin_rf, newdata = qcs_grid)
select(p, X, Y, depth, est, est_non_rf, omega_s) |>
  as_tibble() |> head(n = 2)
```

TODO MOVE TO CAPTION? We have an effect of depth and a spatial random field,
which combined in link space result in our overall prediction. Note the
difference between only the fixed effects (left) and including the spatial
random field (right). TODO DESCRIBE PLOTTING AND/OR INCLUDE

```{r pcod-predict-maps, fig.width=11, fig.asp=0.3, out.width="6in", fig.cap="TODO"}
plot_spatial_map <- function(dat, column, title) {
  ggplot(dat, aes(X, Y, fill = {{ column }})) +
    geom_raster() +
    coord_fixed() +
    theme(legend.position = "bottom") +
    ggtitle(title)
}
g1 <- plot_spatial_map(p, plogis(est_non_rf), "Fixed effects")
g2 <- plot_spatial_map(p, omega_s, "Spatial random field") +  scale_fill_gradient2()
g3 <- plot_spatial_map(p, plogis(est), "Combined prediction")
cowplot::plot_grid(g1, g2, g3, ncol = 3)
```

@@

As a second example, we will construct a spatiotemporal model of catch rates of
Pacific Spiny Dogfish (*Squalus suckleyi*) from a trawl survey off the west
coast of Vancouver Island, Canada. This example extends the spatial model
described above by also including (1) spatiotemporal fields allowing the
distribution in each year to be unique, (2) a time-varying intercept as an
AR(1) process to allow each year to have a different mean catch rate, (3)
a smooth effect of depth, allowing catch rates to vary non-linearly with
depth, and (4) spatial anisotropy allowing spatial correlation to be directionally
dependent. Since catch rates are positive, continuous, and contain zeros, we begin
by specifying the response family as a Tweedie distribution [@tweedie1984] with
a log link. We then compare alternative families, spatiotemporal random field
structures, and the inclusion of anisotropy to illustrate the flexibility of
\pkg{sdmTMB}.

<!-- We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`. -->
<!-- We use a penalized smoother for depth as a main effect via `s()` from the \pkg{mgcv} package. -->
<!-- We specify the family as Tweedie [@tweedie1984] to account for positive continuous density values that also contain zeros. -->
<!-- An alternative would be the `delta_gamma()` family [@aitchison1955] or to model catch weight with an offset for log(area swept) (via the `offset` argument). -->

The dataset includes spatial coordinates, year, dogfish catch weight in kg,
area swept in km, and bottom depth:

```{r dog-head-dat, echo=TRUE}
dat <- select(dogfish, longitude, latitude, year, 
  catch_weight, area_swept, depth)
head(dat, n = 3)
```

We add UTM zone 9 columns, create a log depth column for convenience, and
create a basic mesh:

```{r dog-utms, echo=TRUE}
dat <- add_utm_columns(dat, c("longitude", "latitude"), 
  units = "km", utm_crs = 32609)
dat$log_depth <- log(dat$depth)
mesh <- make_mesh(dat, xy_cols = c("X", "Y"), cutoff = 8)
```

@@

We can then specify our model. We include an offset for the effort variable log
area swept such that we are effectively modelling density and our predictions
will be for an area swept of 1 km^2^.

Our model can be written as TODO

```{r dog-tw, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
fit_tw <- sdmTMB(
  catch_weight ~ s(log_depth),
  data = dat,
  mesh = mesh,
  offset = log(dat$area_swept),
  time = "year",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  spatial = "on",
  spatiotemporal = "iid",
  anisotropy = TRUE,
  silent = FALSE,
  family = tweedie()
)
```

We can explore four alternative families that may better represent the data.
Each alternative family uses a delta or hurdle distribution: delta-gamma and
delta-lognormal (REF), and delta-Poisson-link-gamma and
delta-Poisson-link-lognormal (REF). Note that we choose to omit the
spatiotemporal random fields from the first model component 
(`spatiotemporal = list("off", "iid")`) because initial fits indicated this random field was
collapsing to zero indicating minimal changes in encounter spatial distribution
from year to year.

TODO EQUATIONS

@@

We can then compare the models via AIC:

@@

We find that the Poisson-link-lognormal model is favoured with AIC. We next
test two additional model formulations: making the spatial correlation
isotropic (the default) instead of anisotropic, and structuring the positive
model component spatiotemporal random fields as AR(1) to allow spatiotemporal
hotspots to persist.

The AR(1) fields can be represented as TODO.

@@

<!-- Need description of anisotropy and why it's important here or somewhere
else -->

We find that the anisotropic AR(1) is favoured. It makes sense that anisotropy
is important here given the elongated shape of the continental shelf with
a rapid transition to deeper water. On inspecting the model output with
`summary()`, we notice that the spatial random field SD in the positive model
component has become very small once AR(1) spatiotemporal fields were added. We
therefore consider a model where these are turned off in the second component
of the delta model:

@@

We save our chosen model to the object `fit` to simplify subsequent code, run
the `sanity()` check (suppressed for space), and inspect `summary()`:

@@

The output is considerably more complex compared to our binomial spatial model.
We now have two model components, which are shown one after the other. Starting
with the binomial component, we have output from the smoother, which includes
a linear component (`slog_depth`) and the standard deviation on the smoother
weights (`sds(log_depth)`). The smoother summary follows the format used in the
\pkg{brms} package [@burkner2017]. Next, we have the time-varying intercepts
and information on our anisotropic spatial correlation. We then have the second
model component (lognormal) with a similar summary structure but with the
addition of a dispersion parameter for the lognormal, the AR(1) correlation of
the spatiotemporal random fields, and a spatiotemporal random field marginal
standard deviation. We can use `plot_anisotropy()` to visually inspect the
anisotropy (Figure \ref{fig:dog-aniso}).

```{r dog-aniso, echo=TRUE, fig.cap="TODO", out.width="4in"}
plot_anisotropy(fit_dpl)
```

Similarly to the first example, we can visualize model predictions on a grid
covering the area of interest. Because this is a spatiotemporal model, we first
need to replicate our grid for each year we will predict on. Since this is
a common operation, we include the convenience function `replicate_df()` to
replicate a data frame. We then ensure our data frame contains all the
predictors used in the model (here `log_depth`).

<!-- Propose moving this to a short paragraph `Prediction` to highlight modularity -->
```{r dog-grid, echo=TRUE}
grid <- replicate_df(wcvi_grid, "year", time_values = unique(dat$year))
grid$log_depth <- log(grid$depth)
head(grid, n = 2)
```

@@

@@

Our prediction data frame is similar to the binomial spatial model, but
includes columns for the two delta model components (labelled with suffixes `1`
and `2`) and adds an `epsilon_st` column for spatiotemporal random effects. We can
easily generate plots from this data frame with ggplot2 or similar. We suppress
that code in this paper for brevity.

@@

```{r dog-wvci-pred, fig.asp = 0.5, fig.cap="Example prediction components... TODO", fig.width=9, out.width="6in"}
cowplot::plot_grid(
  g_nonrf,
  g_omega,
  g_eps,
  g_est,
  ncol = 2L
)
```

We can visualize the conditional effect of the bottom depth smoother by
predicting across a sequence of depths and holding other variables at reference
values (Figure \ref{fig:dog-depth-plot}). Here, we pick the last year, specify
to include both delta model components (`model = NA`), omit the random fields
(`re_form = NA`), and return standard errors (`se_fit = TRUE`).

```{r dog-depth, echo=TRUE, out.width="5in"}
nd <- data.frame(
  log_depth = seq(min(dat$log_depth), max(dat$log_depth), length.out = 100), 
  year = max(dat$year)
)
pred_depth <- predict(
  fit, newdata = nd, 
  model = NA, re_form = NA, se_fit = TRUE
)
```

```{r dog-depth-plot, echo=FALSE, fig.cap="TODO", out.width="4in"}
ggplot(pred_depth, aes(
  exp(log_depth), exp(est), 
  ymin = exp(est - 2 * est_se), 
  ymax = exp(est + 2 * est_se))) +
  geom_ribbon(fill = "grey90") +
  geom_line() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(x = "Depth (m)", y = "Density")
```

We can generate an area-weighted population index (e.g., a relative or absolute
index of abundance or biomass) that is independent of sampling locations by
predicting from the model on a grid covering the area of interest and summing
the predicted biomass with the `get_index()` function (Figure
\ref{fig:dog-index-plot}). We supply the grid cell area (4 km^2^) to the `area`
argument and specify `bias_correct = TRUE` to enable a generic epsilon-bias
correction needed due to the non-linear transformation of the random effects
(REF).

@@

```{r dog-index-plot, echo=FALSE, fig.cap="TODO", out.width="4in"}
ggplot(ind, aes(year, est, ymin = lwr, ymax = upr)) +
  geom_pointrange() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.03)), limits = c(0, NA)) +
  labs(y = "Biomass", x = "Year")
```

@@

In our final example, we demonstrate a model with spatially varying coefficient
(SVC) effects and illustrate combining uncertainty from parameters by working
with draws from the joint parameter precision matrix. Spatially varying
coefficients ... TODO REFS. In addition, we illustrate using restricted maximum
likelihood (REML), which integrates over the fixed effects as well as the
random effects.

@@

The spatially varying coefficient effect at any point in space is the
combination of the main effect and spatial random effect for `nao`. Here we
illustrate deriving uncertainty on their combination by simulating from the
parameter precision matrix assuming the joint parameter covariance matrix in
multivariate normal. We do this by specifying a non-null number of simulation
draws to `nsim` in `predict.sdmTMB()`. By default, `nsim > 0` will return
a matrix of draws from the overall prediction. Here, we we instead specify that
we want to return draws from the `zeta_s` random field, which is the spatially
varying random field. This returns a matrix where each row matches a row of
`newdata` and each column is a simulation draw. We then use the function
`spread_sims()` to draw 200 simulations from the parameters themselves, combine
the main effect draws (`sims$nao`) with the random field values `zeta_s`, and
calculate the median for each row of data so we can plot them.

```{r p-mean, message=FALSE, echo=TRUE, eval=TRUE, cache=TRUE}
zeta_s <- predict(fit_owl, newdata = snow, nsim = 200, sims_var = "zeta_s")
dim(zeta_s)
sims <- spread_sims(fit_owl, nsim = 200)
combined <- sims$nao + t(zeta_s)
snow$nao_effect <- apply(exp(combined), 2, median)
```

We can make a basic plot using the following code. A more elaborate version is
shown in Figure \ref{fig:owl-plot-fancy}.

```{r owl-plot-basic, echo=TRUE, eval=FALSE}
ggplot(snow, aes(X, Y)) + geom_point(aes(colour = nao_effect))
```

Overall, we find a weak average positive effect of annual mean NAO on overall
counts, but a southeast to northwest gradient in the intensity of the effect
(Figure \ref{fig:owl-plot-fancy}). This result is consistent with owls closest
to the Atlantic coast and those migrating the furthest south being the most
affected by NAO.

<!-- TODO keep this? -->
<!-- We provide an extended version of this example in Appendix \ref{app:owls}. -->

@@

```{r pcod-eg1-tidy-fe, eval=FALSE, echo=TRUE}
# tidy(fit, conf.int = TRUE)
```

```{r pcod-eg1-tidy-re, eval=FALSE, echo=TRUE}
# tidy(fit, effects = "ran_pars", conf.int = TRUE)
```

<!-- The latter `tidy()` call returns all variance components of the model and omits the standard errors since these are calculated in log space. -->
